From the function-sheaf dictionary to quasicharacters of $p$-adic tori

### 2015-07-21 dictionary.tex
Split Kunneth remark from Z2xZ2 example, remove reference to Bernstein's comment

By any measure, there are more quasicharacter sheaves for $T$ than quasicharacters of $T(K)$.  
In this regard, we are reminded of the work of David Vogan \cite{vogan:93a}, in which he finds a geometrization of complete Langlands parameters for $p$-adic groups 
and, in the process, is led to study the representations of all the pure rational forms of the $p$-adic group, simultaneously.
A similar phenomenon appears in recent work by Joseph Bernstein in which his geometric Ansatz leads to the study of certain sheaves on the stacky classifying space of the $p$-adic group, resulting in a category which appears to be tied to the representations of all the pure rational forms of the $p$-adic group \cite{bernstein:vogan_conference}.
Indeed, Bernstein has suggested to us that our category of quasicharacters for $T$ may be tied to quasicharacters of all the pure rational forms of $T$; we do not pursue that suggestion in this paper.

\medskip
\noindent\textbf{Relation to other work.\ }
The main use of the term character sheaf is of course due to Lusztig.
It is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on certain reductive groups over algebraically closed fields in the series of papers
beginning with \cite{lusztig:disconnected1}.
When commutative, it is not difficult to relate Frobenius-stable character sheaves to our character sheaves (Remark~\ref{rem:Lusztig}). The new features that we have found pertaining to Weil sheaves and the non-triviality of $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$ do not arise in that context.

@@

\begin{remark} \label{eg:H2}
Recall the K\"unneth formula in group cohomology \cite{brown:CohomologyOfGroups}*{Prop. I.0.8}: if $A$ and $A'$ are groups and $M$ and $M'$ are abelian groups with $M$ $\ZZ$-free, then
\[
\Hh^n(A \times A', M \otimes M') \cong \bigoplus_{i+j=n} \Hh^i(A, M) \otimes \Hh^j(A', M') \ \ \oplus \bigoplus_{i+j=n+1} \Tor_1^{\ZZ}\left(\Hh^i(A, M), \Hh^j(A', M')\right).
\]

Now suppose $\bG = \ZZ^r \times \prod_{i = 1}^m \ZZ/N_i\ZZ$ is an arbitrary finitely generated abelian group,
with $N_i \mid N_{i+1}$.  Then the K\"unneth formula implies that
\begin{equation} \label{eq:H2comp}
\Hh^2(\bG, \EEx) \cong \left(\EEx\right)^{r(r-1)/2} \times \prod_{i=1}^m (\ZZ/N_i\ZZ)^{m+r-i}.
\end{equation}
We see that $\Hh^2(\bG, \EEx)$ is trivial if and only if $\bG$ is cyclic.  Of course, $\Hh^0(\Weil{}, \Hh^2(\bG, \EEx))$
may or may not be trivial, even when $\Hh^2(\bG, \EEx)$ is non-trivial.

Consider the simplest non-trivial case, where $\bG = \{1, i, j, k\} \cong \ZZ/2\ZZ \times \ZZ/2\ZZ.$  Using eqref{eq:H2comp},
we have $\Hh^2(\bG, \EEx) \cong \ZZ/2\ZZ$, on which $\Weil{}$ must act trivially, regardless of its action on $\bG$ itself.
The non-trivial element corresponds to the extension
\begin{equation} \label{eq:Qdef}
1 \to \EEx \to Q \to \bG \to 1,
\end{equation}
where $Q = \{c + c_ii + c_jj + c_kk \tq c,c_i,c_j,c_k \in \EE \mbox{ with exactly one nonzero}\}$ is a subgroup of the quaternion algebra over $\EE$.
Let $a$ be a $2$-cocycle corresponding to this extension, with values in $\{\pm 1\}$. When $\Frob{G}$ acts trivially on $\bG$,
any homomorphism $b : \bG \to \EEx$ will satisfy \eqref{eq:nohom}, and the corresponding $\alpha \oplus \beta$ are non-cohomologous in $\Hh^2(E_G^\bullet)$.
When $\Frob{G}$ exchanges $i$ and $j$, then we may take $b(1) = 1, b(i) = -1$ and $b(j) = b(k) = \pm 1$, up to coboundaries.
Finally, when $\Frob{G}$ cycles $i$, $j$ and $k$, any homomorphism $b : \bG \to \EEx$ will satisfy \eqref{eq:nohom},
but now the corresponding $\alpha \oplus \beta$ are all cohomologous in $\Hh^2(E_G^\bullet)$.  In each case, we may produce
an explicit character sheaf from the listed $a$ and $b$.

Note that these character sheaves arise from discrete isogenies, as in Section \ref{ssec:discrete_isogenies}.
Let $\bH$ be the quaternion group of order $8$: the subgroup of $Q$ with $c, c_i, c_j, c_k \in \{\pm 1\}$.
The sequence \eqref{eq:Qdef} is the pushforward of
\[
1 \to \{\pm 1\} \to \bH \to \bG \to 1
\]
along the inclusion $\{\pm 1\} \hookrightarrow \EEx$.  Note that these character sheaves arise from a non-commutative cover of $\bG$,
justifying the inclusion of such covers in the definition of a discrete isogeny.
\end{remark}

>>>>

\iffalse
By any measure, there are more quasicharacter sheaves for $T$ than quasicharacters of $T(K)$.  
In this regard, we are reminded of the work of David Vogan \cite{vogan:93a}, in which he finds a geometrization of complete Langlands parameters for $p$-adic groups 
and, in the process, is led to study the representations of all the pure rational forms of the $p$-adic group, simultaneously.
A similar phenomenon appears in recent work by Joseph Bernstein in which his geometric Ansatz leads to the study of certain sheaves on the stacky classifying space of the $p$-adic group, resulting in a category which appears to be tied to the representations of all the pure rational forms of the $p$-adic group \cite{bernstein:vogan_conference}.
Indeed, Bernstein has suggested to us that our category of quasicharacters for $T$ may be tied to quasicharacters of all the pure rational forms of $T$; we do not pursue that suggestion in this paper.
\fi

\medskip
\noindent\textbf{Relation to other work.\ }
The main use of the term character sheaf is of course due to Lusztig.
It is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on certain reductive groups over algebraically closed fields in the series of papers
beginning with \cite{lusztig:disconnected1}.
When commutative, it is not difficult to relate Frobenius-stable character sheaves to our character sheaves (Remark~\ref{rem:Lusztig}). The new features that we have found pertaining to Weil sheaves and $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$ do not arise in that context because, for such groups, Weil sheaves are unnecessary (\S \ref{ssec:alg_groups} and $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}=0$ (Remark \ref{rem:H2}).

@@

\begin{remark}
Recall the K\"unneth formula in group cohomology \cite{brown:CohomologyOfGroups}*{Prop. I.0.8}: if $A$ and $A'$ are groups and $M$ and $M'$ are abelian groups with $M$ $\ZZ$-free, then
\[
\Hh^n(A \times A', M \otimes M') \cong \bigoplus_{i+j=n} \Hh^i(A, M) \otimes \Hh^j(A', M') \ \ \oplus \bigoplus_{i+j=n+1} \Tor_1^{\ZZ}\left(\Hh^i(A, M), \Hh^j(A', M')\right).
\]
%
Now suppose $\bG = \ZZ^r \times \prod_{i = 1}^m \ZZ/N_i\ZZ$ is an arbitrary finitely generated abelian group,
with $N_i \mid N_{i+1}$.  Then the K\"unneth formula implies that
\begin{equation} \label{eq:H2comp}
\Hh^2(\bG, \EEx) \cong \left(\EEx\right)^{r(r-1)/2} \times \prod_{i=1}^m (\ZZ/N_i\ZZ)^{m+r-i}.
\end{equation}
We see that $\Hh^2(\bG, \EEx)$ is trivial if and only if $\bG$ is cyclic.  Of course, $\Hh^0(\Weil{}, \Hh^2(\bG, \EEx))$
may or may not be trivial, even when $\Hh^2(\bG, \EEx)$ is non-trivial.
\end{remark}

\begin{example}\ref{eg:H2}
Consider the simplest non-trivial case, where $\bG = \{1, i, j, k\} \cong \ZZ/2\ZZ \times \ZZ/2\ZZ.$  Using eqref{eq:H2comp},
we have $\Hh^2(\bG, \EEx) \cong \ZZ/2\ZZ$, on which $\Weil{}$ must act trivially, regardless of its action on $\bG$ itself.
The non-trivial element corresponds to the extension
\begin{equation} \label{eq:Qdef}
1 \to \EEx \to Q \to \bG \to 1,
\end{equation}
where $Q = \{c + c_ii + c_jj + c_kk \tq c,c_i,c_j,c_k \in \EE \mbox{ with exactly one nonzero}\}$ is a subgroup of the quaternion algebra over $\EE$.
Let $a$ be a $2$-cocycle corresponding to this extension, with values in $\{\pm 1\}$. When $\Frob{G}$ acts trivially on $\bG$,
any homomorphism $b : \bG \to \EEx$ will satisfy \eqref{eq:nohom}, and the corresponding $\alpha \oplus \beta$ are non-cohomologous in $\Hh^2(E_G^\bullet)$.
When $\Frob{G}$ exchanges $i$ and $j$, then we may take $b(1) = 1, b(i) = -1$ and $b(j) = b(k) = \pm 1$, up to coboundaries.
Finally, when $\Frob{G}$ cycles $i$, $j$ and $k$, any homomorphism $b : \bG \to \EEx$ will satisfy \eqref{eq:nohom},
but now the corresponding $\alpha \oplus \beta$ are all cohomologous in $\Hh^2(E_G^\bullet)$.  In each case, we may produce
an explicit character sheaf from the listed $a$ and $b$.

Note that these character sheaves arise from discrete isogenies, as in Section \ref{ssec:discrete_isogenies}.
Let $\bH$ be the quaternion group of order $8$: the subgroup of $Q$ with $c, c_i, c_j, c_k \in \{\pm 1\}$.
The sequence \eqref{eq:Qdef} is the pushforward of
\[
1 \to \{\pm 1\} \to \bH \to \bG \to 1
\]
along the inclusion $\{\pm 1\} \hookrightarrow \EEx$.  Note that these character sheaves arise from a non-commutative cover of $\bG$,
justifying the inclusion of such covers in the definition of a discrete isogeny.
\end{example}
